Do we need detailed demographic data to forecast plant population responses to climate change?

### 2016-08-30 manuscript/manuscript.Rmd
adds PBA line edits for R1; ready for USGS

\begin{enumerate}[label=\textbf{\arabic*}]
        \item Rapid environmental change has generated growing interest in forecasts of future population trajectories. Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions. An alternative is to build models using population-level data that are much easier to collect than individual-level data over broad spatial scales. However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
        \item Here, we test the consequences of aggregating individual responses when forecasting the population states \textcolor{blue}{(percent cover)} and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA. We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting \textcolor{blue}{accuracy} and forecast horizons with and without the inclusion of climate covariates. For both models, we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction. 
        \item In the absence of climate effects, we found no significant difference between the forecast \textcolor{blue}{accuracy} of models based on individual-level data and models based on population-level data. Climate effects were weak, but increased forecast \textcolor{blue}{accuracy} for two species. Increases in \textcolor{blue}{accuracy} with climate covariates were similar between model types.
        \item In our case study, percent cover models generated forecasts as \textcolor{blue}{accurate} as those from a demographic model. For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models. Long time series of percent cover data already exist for many plant species. Modelers should exploit these data to predict the impacts of environmental change.
    \end{enumerate}

@@

To date, most empirical population modelers have relied on individual-level data, with few attempts to capitalize on population-level measures. An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control. They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure. Such models do not require individual-level demographic data and can adequately describe population dynamics. Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales, even if some precision is lost when compared to fully continuous models. However, previous tests of density-structured population models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.

<!--
\textcolor{blue}{Density-structured models based on animal counts have also begun to find a home in dynamic species distribution models} [@Mieszkowska2013; @Hefley2016].
-->

Addressing climate change questions with models fit to population-level data is potentially problematic. Population growth (or decline) is the outcome of demographic processes such as survival, growth, and recruitment that occur at the level of individual plants. Climate can affect each demographic process in unique, potentially opposing, ways [@Dalgleish2011]. These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable. Futhermore, models based on aggregated data may reflect short-term effects of one vital rate more than others whose importance may only emerge over the long-term. For example, a one-year change in a plant species' cover or biomass might reflect individual growth or shrinkage, whereas the long-term trajectory of the population might be more influenced by recruitment. The same is true for density dependence: intraspecific density depedence may act most strongly on vital rates, like recruitment, that are difficult to identify from population-level data. If density dependence and/or important climate effects are missed because of the aggregation inherent in population-level data, then population models built with such data will make uninformative or unreliable forecasts.

@@

We fit all models using a hierarchical Bayesian approach. We focus on the main process and the model likelihood in what follows (full model descriptions are in the Supporting Information). For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P* for survival, growth, recruitment, or population growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not, for all observation years and quadrats. All model parameters are species-specific, but we omit subscripts for species in model descriptions below to reduce visual clutter. For brevity, we only describe models with climate covariates included, but models without climate covariates are simply the models described below with the climate effects removed.

@@

### Model fitting and statistical regularization
####  Model fitting
Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical. For each species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were modeled with global distributions representing the mean size effect and intercept. Quadrat random effects were also fit hierarchically, with quadrat offsets modeled with distributions with mean zero and a shared variance term (independent Gaussian priors). Climate effects were modeled as independent covariates whose prior distributions were optimized for prediction using statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).

@@

### Forecast horizons
An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases.
\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon} [@Petchey2015].
To assess the forecast horizons of our models, we initiate the forecast model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
For example, if we initialize the forecast model with percent cover in 1940, we are able to make five forecasts up to the year 1945.
Forecast models are not re-initialized with observations between years.
Thus, in our current example, the model forecast for percent cover in 1941 has a forecast horizon of one year, the forecast in 1942 has a forecast horizon of two years, and so on.
We performed these simulations for all model types (IPM with/without climate; QBM with/without climate) using mean parameter values for all possible initial years.
For a given forecast distance, we averaged the correlation between forecasts and observations.
Note that our forecasts for the horizon analysis are all made using in-sample data because we used model fits from the full data set.
Nonetheless, our simulations offer insight into the differences among model forecast horizons.
\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold.
For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes} [@Petchey2015], \textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25% of the variance in observations is explained by the predictions.
So, the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.}

@@

With climate covariates included \textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased}, but they did so at similar rates (Fig. 4).
The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the \textcolor{blue}{temporal distance of the forecast increased}, whereas IPM forecast accuracy declined (Fig. 4).
\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H. comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast profieciency threshold by a large margin (Fig. 4, Table S24).}

@@

Our comparison of a traditional, demographic population model without environmental forcing (the IPM) to an equivalent model inspired by density-structured models (the QBM) showed that, generally, IPM forecasts of out-of-sample plant population states were no more accurate than forecasts from the QBM (Fig. 2; `no-climate' bars).
This result differed from our expectation that the IPM would out-perform the QBM, because of its mechanistic representation of the perennial life cycle.
Our result also confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.

We also expected the inclusion of environmental forcing to reveal more differences between the models.
Interannual variation in weather can affect vital rates in different ways [@Dalgleish2011].
Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
We found some evidence that the QBM failed to detect climate effects for three species (*B. gracilis*, *H. comata*, and *Poa secunda*), where important climate effects were identified in the individual vital rate models but not in the percent cover model (Fig. 3).
However, for the two species where including climate covariates increased forecast accuracy (*B. gracilis* and *Poa secunda*), forecast skill (Fig. 2) and error (Fig. S2) were equivalent between the IPM and QBM.

@@

We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the \textcolor{blue}{time between the model initialization and the forecast increased.}
In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
Over time, the addition and subtraction of small plants can have large effects on population growth, thus, explicitly modeling these effects could contribute to a longer forecast horizon. 
However, we found no evidence \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
Thus, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}

@@

\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species. 
Such short forecast horizons are not encouraging.}
\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed} [@Petchey2015; @Mouquet2015].
\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate} [@Teller2016], \textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty} [@Petchey2015].
\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
\textcolor{blue}{Whether the poor predictive ability of our models impacts their comparison relative to one another is an open question}.

In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data.
This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
Moreover, for the two species where including climate covariates improved forecast skill, both models generated equally skillful one-step-ahead forecasts.
Thus, we conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the states and dynamics of plant populations.
Unfortunately, our analysis, where climate effects were relatively unimportant in vital rate regressions, did not allow us to sufficiently test our prediction that individual-level data is necessary to generate skillful forecasts if different vital rates respond to climate in unique, potentially opposing, ways.
Nonetheless, our results are encouraging for the use of easy-to-collect population-level data for forecasting the state of plant populations.

>>>>

\begin{enumerate}[label=\textbf{\arabic*}]
        \item Rapid environmental change has generated growing interest in forecasts of future population trajectories. Traditional population models built with detailed demographic observations from one study site can address the impacts of environmental change at particular locations, but are difficult to scale up to the landscape and regional scales relevant to management decisions. An alternative is to build models using population-level data that are much easier to collect over broad spatial scales than individual-level data.  However, it is unknown whether models built using population-level data adequately capture the effects of density-dependence and environmental forcing that are necessary to generate skillful forecasts.
        \item Here, we test the consequences of aggregating individual responses when forecasting the population states \textcolor{blue}{(percent cover)} and trajectories of four perennial grass species in a semi-arid grassland in Montana, USA. We parameterized two population models for each species, one based on individual-level data (survival, growth and recruitment) and one on population-level data (percent cover), and compared their forecasting \textcolor{blue}{accuracy} and forecast horizons with and without the inclusion of climate covariates. For both models, we used Bayesian ridge regression to weight the influence of climate covariates for optimal prediction. 
        \item In the absence of climate effects, we found no significant difference between the forecast \textcolor{blue}{accuracy} of models based on individual-level data and models based on population-level data. Climate effects were weak, but increased forecast \textcolor{blue}{accuracy} for two species. Increases in \textcolor{blue}{accuracy} with climate covariates were similar between model types.
        \item In our case study, percent cover models generated forecasts as \textcolor{blue}{accurate} as those from a demographic model. For the goal of forecasting, models based on aggregated individual-level data may offer a practical alternative to data-intensive demographic models. Long time series of percent cover data already exist for many plant species. Modelers should exploit these data to predict the impacts of environmental change.
    \end{enumerate}

@@

To date, most empirical population modelers have relied on individual-level data, with few attempts to capitalize on population-level measures. An important exception was an effort by @Taylor2004 to model the population growth rate of an invasive species to identify the best strategies for invasion control. They used a "density-structured" model where the state variable is a discrete density state rather than a continuous density measure. Such models do not require individual-level demographic data and can adequately describe population dynamics. Building on @Taylor2004, @Freckleton2011 showed that density-structured models compare well to continuous models in theory, and @Queenborough2011 provide empirical evidence that density-structured models are capable of reproducing population dynamics at landscape spatial scales [also see @Mieszkowska2013], even if some precision is lost when compared to fully continuous models. However, previous tests of density-structured population models have yet to assess their ability to forecast out-of-sample observations, and they have not included environmental covariates, which are necessary to forecast population responses to climate change.

Addressing climate change questions with models fit to population-level data is potentially problematic. Population growth (or decline) is the outcome of demographic processes such as survival, growth, and recruitment that occur at the level of individual plants. Climate can affect each demographic process in unique, potentially opposing, ways [@Dalgleish2011]. These unique climate responses may be difficult to resolve in statistical models based on population-level data where demographic processes are not identifiable. Futhermore, models based on aggregated data may reflect short-term effects of one vital rate more than others whose importance may only emerge over the long-term. For example, a one-year change in a plant species' cover or biomass might reflect growth or shrinkage of the largest individuals, whereas the long-term trajectory of the population might be more influenced by recruitment. The same is true for density dependence: intraspecific density depedence may act most strongly on vital rates, like recruitment, that are difficult to identify from population-level data. If density dependence and/or important climate effects are missed because of the aggregation inherent in population-level data, then population models built with such data will make uninformative or unreliable forecasts.

@@

We fit all models using a hierarchical Bayesian approach. In the following description, we focus on the main process and the model likelihood (full model descriptions are in the Supporting Information). For the likelihood models, $\textbf{y}^X$ is always the relevant vector of observations for vital rate *X* (*X* = *S*, *G*, *R*, or *P* for survival, growth, recruitment, or population growth). For example, $\textbf{y}^S$ is a vector of 0s and 1s indicating whether a genet survives from *t* to *t+1*, or not, for all observation years and quadrats. All model parameters are species-specific, but we omit subscripts for species in model descriptions below to reduce visual clutter. For brevity, we only describe models with climate covariates included, but models without climate covariates are simply the models described below with the climate effects removed.

@@

### Model fitting and statistical regularization
####  Model fitting
Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those priors should be hierarchical. For each species, we fit yearly size effects and yearly intercepts hierarchically, where year-specific coefficients were modeled with global distributions representing the mean size effect and intercept. Quadrat random effects were also fit hierarchically, with quadrat offsets modeled using distributions with mean zero and a shared variance term (independent Gaussian priors). Climate effects were modeled as independent covariates whose prior distributions were optimized for prediction using statistical regularization (see **Statistical regularization: Bayesian ridge regression** below).

@@

### Forecast horizons
An important feature of any forecasting model is the rate at which forecast skill declines as the time between an observation and a forecast increases.
\textcolor{blue}{In particular, we are interested in the temporal distance at which forecast skill falls below a threshold: the so-called ecological forecast horizon} [@Petchey2015].
To assess the forecast horizons of our models, we initiate the forecast model with the population state at some time *t* and make sequential forecasts of the population at times $t+1, t+2, \dots, t+T$ where *T* is the maximum number of years between the initial year and the final year of our observations.
For example, if we initialize the forecast model with percent cover in 1940, we are able to make five forecasts up to the year 1945.
Forecast models are not re-initialized with observations between years.
Thus, in our current example, the model forecast for percent cover in 1941 has a forecast horizon of one year, the forecast in 1942 has a forecast horizon of two years, and so on.
We performed these simulations using mean parameter values for all model types (IPM with/without climate; QBM with/without climate) and all possible initial years.
For a given forecast distance, we averaged the correlation between forecasts and observations.
Note that our forecasts for the horizon analysis are all made using in-sample data because we used model fits from the full data set.
Nonetheless, our simulations offer insight into the differences among model forecast horizons.
\textcolor{blue}{We chose an arbitrary forecast accuracy of $\rho = 0.5$ as our forecast proficiency threshold. the forecast horizon is the temporal distance at which forecast accuracy falls below $\rho = 0.5$.
For basic research on forecasting, arbitrary proficiency thresholds suffice for comparative purposes} [@Petchey2015], \textcolor{blue}{and $\rho = 0.5$ represents the point at which about 25\% of the variance in observations is explained by the predictions.}

@@

With climate covariates included \textcolor{blue}{and using mean parameter values, the accuracy of both models' forecasts declined as the distance between the last observation and the forecast increased}, but they did so at similar rates (Fig. 4).
The only exception is for *Poa secunda*, where QBM forecast accuracy remained steady as the \textcolor{blue}{temporal distance of the forecast increased}, whereas IPM forecast accuracy declined (Fig. 4).
\textcolor{blue}{The forecast horizons were short: forecast accuracy fell below $\rho$ = 0.5 after one year for the IPM for most species, and after four years, at most, for the QBM (Fig. 4).
Across the different temporal distances from the observation to the forecast, the IPM was never more accurate than the QBM (\emph{P} > 0.05 for all one-sided \emph{t}-tests, Table S23).
Likewise, the QBM was rarely more accurate the IPM, the only exception being for \emph{H. comata} at temporal distances of two ($t_{(115)}$ = 2.39, \emph{P} = 0.002) and three years ($t_{(98)}$ = 2.04, \emph{P} = 0.022) (Table S24).
There were some cases where the QBM was more accurate than the IPM for \emph{Poa secunda}, but neither model exceeded the forecast proficiency threshold by a large margin (Fig. 4, Table S24).}

@@

Our comparison between a traditional, demographic population model without environmental forcing (the IPM) and an equivalent model inspired by density-structured models (the QBM) showed that IPM forecasts of out-of-sample plant population states were no more accurate than forecasts from the QBM (Fig. 2; `no-climate' bars).
This result differed from our expectation that the IPM would out-perform the QBM, because of its mechanistic representation of the perennial life cycle.
Our result also confirms theoretical [@Freckleton2011] and empirical work [@Taylor2004; @Queenborough2011] showing that density-structured models can be useful surrogates for demographic models when the goal is to estimate or forecast population states over large spatial extents.

We also expected the inclusion of environmental forcing to reveal further differences between the models.
Interannual variation in weather can affect vital rates in different ways [@Dalgleish2011].
Thus, estimates of climate effects on plant population growth may be biased or non-identifiable when the underlying statistical model is fit using population-level data that integrates over the potentially unique climate responses of individual vital rates.
We found some evidence that the QBM failed to detect climate effects for three species (*B. gracilis*, *H. comata*, and *Poa secunda*), where important climate effects were identified in the individual vital rate models but not in the percent cover model (Fig. 3).
\textcolor{blue}{For \emph{H. comata}, adding climate covariates did not improve forecasts (Fig. 2), despite the significant climate effects in the vital rate regressions (Fig. 3).}
Furthermore, for the two species where including climate covariates increased forecast accuracy (*B. gracilis* and *Poa secunda*), forecast accuracy (Fig. 2) and error (Fig. S2) were equivalent between the IPM and QBM.

@@

We also expected IPM forecast accuracy to decline at a lower rate than the QBM as the \textcolor{blue}{time between the model initialization and the forecast increased.}
In principle, more mechanistic models should produce better predictions, especially under novel conditions [@Evans2012b; @Schindler2015b].
In our case, the IPM explicitly models the influence of weather on recruitment and survival, effects that may be poorly represented in the QBM because recruitment and survival mainly affect small plants that contribute little to year-to-year changes in percent cover.
Over longer time scales, the addition and subtraction of small plants could have large effects on population growth, so explicitly modeling these effects could contribute to a longer forecast horizon. 
However, we found no evidence \textcolor{blue}{that the forecast horizon for the IPM was greater than the QBM (Fig. 4).
On the contrary, the QBM tended to have a slightly longer forecast horizon than the IPM for most species (Fig. 4).
The QBM has fewer processes and parameters, which can reduce bias due to parameter uncertainty.
As a result, the QBM may better capture near term dynamics when populations do not fluctuate widely, as in our case.}

@@

\textcolor{blue}{Although our main goal was to compare individual-based and population-level modeling approaches relative to one another, it is worth reflecting on the absolute forecasting skill of our models.
In particular, the forecast horizon of both models, defined as the time horizon at which the correlation between predictions and observations falls below $\rho = 0.5$, is less than five years for all species. 
Such short forecast horizons are not encouraging.}
\textcolor{blue}{Unfortunately, we have few ideas about how to improve population forecasts that have not already been proposed} [@Petchey2015; @Mouquet2015].
\textcolor{blue}{Longer time-series should improve our ability to detect exogenous drivers such as climate} [@Teller2016], \textcolor{blue}{and modeling larger spatial extents may reduce parameter uncertainty} [@Petchey2015].
\textcolor{blue}{We may also have to shift our perspective from making explicit point forecasts to making moving average forecasts} [@Petchey2015].
\textcolor{blue}{Whether the poor predictive ability of our models impacts the comparison of models based on individual vs. population-level data is an open question}.

In conclusion, we found that models based on individual-level demographic data generally failed to generate more skillful population forecasts than models based on population-level data, even in models which included climate covariates.
This finding runs counter to our expectations, but is consistent with recent theoretical [@Freckleton2011] and empirical work [@Queenborough2011].
We conclude that models based on population-level data, rather than individual-level data, may be adequate for forecasting the states and dynamics of plant populations.
This conclusion comes with the caveat that our analysis may be a weak test of the prediction that individual-level data is necessary for forecasting if different vital rates respond to climate in opposing ways, because climate effects were relatively unimportant in our vital rate regressions.
Nonetheless, our results should encourage the use of easy-to-collect population-level data for forecasting the state of plant populations.
